Open archaeology, open source?

### 2024-07-13 analysis/_03-data_methodology.qmd
Correct in-text stats
Also made some small wording changes, and updated the tinytable syntax for links

```{r tbl-forges}
#| tbl-cap: Software forges used by open archaeology projects, as per open-archaeo
oarch_forges |>
  mutate(
    repo_host = recode(
      repo_host,
      "GitHub" = "[GitHub](https://github.com/)",
      "Codeberg" = "[Codeberg](https://codeberg.org/)",
      "GitLab" = "[GitLab](https://about.gitlab.com/)",
      "BitBucket" = "[Bitbucket](https://bitbucket.org/product/)",
      "Launchpad" = "[Launchpad](https://launchpad.net/)"
    ),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  rename(Host = repo_host, `%` = p) |>
  tt() |>
  format_tt(1, markdown = TRUE)
```

Where applicable, we obtained more detailed information about each repository's contents and contribution histories from the GitHub API (application programming interface).
Our analysis incorporates data on `r n_repos` repositories^[We excluded three GitHub repositories from the analysis for technical reasons. For example, https://github.com/carpentries-incubator/R-archaeology-lesson is a repository within the scope of open-archaeo, but it was forked from a pre-existing repository that is not (https://github.com/datacarpentry/R-ecology-lesson) and thus includes in its commit history irrelevant data from the parent repository.], comprising `r n_commits` commits and `r n_issues` issues/pull requests from `r n_contributors` distinct users, as well as repository metadata on programming languages used, licensing, stars and forks, and so on.

@@

That being said, we cannot account for practices that occur through offline or private channels, or forms of collaboration we do not know about. 
We did not directly observe or interview archaeological software developers, though our conclusions do draw heavily from our experience as members of that community ourselves.^[JR is an extremely prolific archaeological open source software developer, who actively builds code, teaches programming methods, and conducts code reviews. Through his maintenance of open-archaeo over the past 5+ years, ZB has honed an extensive understanding of the people who create archaeological software and the institutional ecology that supports their work.] 
Our earliest data is from 2005 and our study can say little about collaborative software development in archaeology before this point, though we know there was a significant amount of it [@ducke2013; @whallon1972].

>>>>

```{r tbl-forges}
#| tbl-cap: Software forges used by open archaeology projects
oarch_forges |>
  mutate(
    repo_host = recode(
      repo_host,
      "GitHub" = "[GitHub](https://github.com/)",
      "Codeberg" = "[Codeberg](https://codeberg.org/)",
      "GitLab" = "[GitLab](https://about.gitlab.com/)",
      "BitBucket" = "[Bitbucket](https://bitbucket.org/product/)",
      "Launchpad" = "[Launchpad](https://launchpad.net/)"
    ),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  rename(Host = repo_host, `%` = p) |>
  tt() |>
  format_tt(j = 1, markdown = TRUE)
```

Where applicable, we obtained more detailed information about each repository's contents and contribution histories from the GitHub API (application programming interface).
Our analysis incorporates data on `r n_repos` repositories^[We excluded three GitHub repositories from the analysis for technical reasons. For example, https://github.com/carpentries-incubator/R-archaeology-lesson is a repository within the scope of open-archaeo, but it was forked from a pre-existing repository that is not (https://github.com/datacarpentry/R-ecology-lesson) and thus includes in its commit history irrelevant data from the parent repository.], `r n_commits` commits, `r n_issues` issues/pull requests, and `r n_comments` from `r n_contributors` distinct users, as well as repository metadata on programming languages used, licensing, stars and forks, and so on.

@@

That being said, we cannot account for practices that occur through offline or private channels, or forms of collaboration we do not know about. 
We did not directly observe or interview archaeological software developers, though our conclusions do draw heavily from our experience as members of that community ourselves.^[JR has been active in the development of open source research software for archaeology for about ten years. Through his maintenance of open-archaeo over the past 5+ years, ZB has honed an extensive understanding of the people who create archaeological software and the institutional ecology that supports their work.] 
Our earliest data is from 2005 and our study can say little about collaborative software development in archaeology before this point, though we know there was a significant amount of it [@ducke2013; @whallon1972].

### 2024-07-13 analysis/_04-open_archaeology.qmd
Correct in-text stats
Also made some small wording changes, and updated the tinytable syntax for links

```{r tbl-categories}
#| tbl-cap: Categories of open archaeology projects
tribble(
  ~category, ~url,                               ~kind,       ~scope,
  "Packages and libraries", "https://open-archaeo.info/categories/packages-and-libraries/",                "Software", "Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.",
  "Standalone software", "https://open-archaeo.info/categories/standalone-software/",                   "Software", "Software that may be operated without needing to first access an underlying platform.",
  "Scripts", "https://open-archaeo.info/categories/scripts/",                               "Software", "Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation.",
  "Specifications, protocols and schemas", "https://open-archaeo.info/categories/specifications-protocols-and-schemas/", "Documents", "A formal data structure or framework intended to be used as a model.",
  "Products", "https://open-archaeo.info/categories/products/",                              "Documents", "Stable outcomes of creative work.",
  "Guides", "https://open-archaeo.info/categories/guides/",                                "Documents", "An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.",
  "Lists and datasets", "https://open-archaeo.info/categories/lists-and-datasets/",                    "Documents", "A series of consistently organized observations assembled with purpose."
) |>
  left_join(oarch, by = "category") |>
  count(kind, category, url, scope) |>
  mutate(
    category = glue("[{category}]({url})"),
    p = n / sum(n),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  arrange(desc(kind), -n) |>
  select(category, scope, n, p) |>
  rename_with(tools::toTitleCase) |>
  rename(`%` = p) |>
  tt() |>
  group_tt(list("Software" = 1, "Documents" = 4)) |>
  format_tt(1, markdown = TRUE)
```

@@

n_packages <- nrow(filter(oarch, category == "Packages and libraries"))
p_packages <- n_packages / nrow(oarch)

@@

n_datasets <- nrow(filter(oarch, category == "Lists and datasets"))
p_datasets <- n_datasets / nrow(oarch)

# n_platform <- sum(!is.na(oarch$platform))
# p_platform <- n_platform / nrow(oarch)

# n_datasets <- nrow(filter(oarch, tag1 == "Datasets" | tag2 == "Datasets" | 
#                             tag3 == "Datasets" | tag4 == "Datasets" | 
#                             tag5 == "Datasets"))
# p_datasets <- n_datasets / nrow(oarch)
```

From our breakdown of open-archaeo by category (see @tbl-categories), we can infer the pervasiveness of various development models, and the requisite technical capabilities that developers assume users hold.
Most resources (`r percent(p_packages)`) included in open-archaeo are designed to be used atop an existing "platform" -- for example a package that extends a programming language (e.g. radiocarbon calibration is implemented in R in the package 'rcarbon' or Python in the package 'iosacal') or a plugin for an application (e.g. 'ArchaeoAstroInsight' adds tools for measuring astronomical alignments to QGIS).
Essentially such projects create additional functions within the base platform that are useful for archaeological purposes.
Others create standalone software (`r percent(p_standalone)`) that can be run independently of such platforms, for example desktop or web apps.
A significant number of projects also comprise of datasets (`r percent(p_datasets)`) and non-packaged code snippets (`r percent(p_scripts)`) that have been made available for general use.

```{r tbl-platforms}
#| tbl-cap: Platforms and programming languages used by open archaeology projects
oarch |>
  drop_na(platform) |>
  count(platform) |>
  mutate(
    significant = n > 1,
    platform = if_else(significant, platform, "Other")
  ) |>
  group_by(platform) |>
  summarise(n = sum(n)) |>
  mutate(
    p = n / sum(n),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  arrange(platform == "Other", -n) |>
  rename_with(tools::toTitleCase) |>
  tt()
```
```{r data-platforms}
  p_platform_r <- sum(oarch$platform == "R", na.rm = TRUE) / sum(!is.na(oarch$platform))
  p_platform_python <- sum(oarch$platform == "Python", na.rm = TRUE) / sum(!is.na(oarch$platform))
  p_platform_qgis <- sum(oarch$platform == "QGIS", na.rm = TRUE) / sum(!is.na(oarch$platform))
  ```

@@

```{r tbl-tags}
#| tbl-cap: Themes of open archaeology projects. See Batist and Roe [-@batist2023 table 1] for a description of each tag's scope.
oarch |>
  pivot_longer(c(tag1, tag2, tag3, tag4, tag5), values_to = "tag",
               values_drop_na = TRUE) |>
  drop_na(tag) |>
  group_by(tag, category) |>
  summarise(n = n(), .groups = "drop_last") |>
  mutate(total = sum(n)) |>
  pivot_wider(names_from = category, values_from = n) |>
  mutate(
    tag = glue("[{tag}](https://open-archaeo.info/tags/{slugify(tag)})"),
    across(everything(), \(x) replace_na(x, 0))
  ) |>
  arrange(-total) |>
  rename(theme = tag) |>
  rename_with(tools::toTitleCase) |>
  tt() |>
  format_tt(1, markdown = TRUE)
```

@@

p_license_gpl <- n_filter(oarch, str_detect(license, "GPL")) / sum(!is.na(oarch$license))
p_license_mit <- n_filter(oarch, str_detect(license, "MIT")) / sum(!is.na(oarch$license))
p_license_cc <- n_filter(oarch, str_detect(license, "CC")) / sum(!is.na(oarch$license))
```

>>>>

```{r tbl-categories}
#| tbl-cap: Categories of open archaeology projects
tribble(
  ~category, ~url,                               ~kind,       ~scope,
  "Packages and libraries", "https://open-archaeo.info/categories/packages-and-libraries/",                "Software", "Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.",
  "Standalone software", "https://open-archaeo.info/categories/standalone-software/",                   "Software", "Software that may be operated without needing to first access an underlying platform.",
  "Scripts", "https://open-archaeo.info/categories/scripts/",                               "Software", "Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation.",
  "Specifications, protocols and schemas", "https://open-archaeo.info/categories/specifications-protocols-and-schemas/", "Documents", "A formal data structure or framework intended to be used as a model.",
  "Products", "https://open-archaeo.info/categories/products/",                              "Documents", "Stable outcomes of creative work.",
  "Guides", "https://open-archaeo.info/categories/guides/",                                "Documents", "An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.",
  "Lists and datasets", "https://open-archaeo.info/categories/lists-and-datasets/",                    "Documents", "A series of consistently organized observations assembled with purpose."
) |>
  left_join(oarch, by = "category") |>
  count(kind, category, url, scope) |>
  mutate(
    category = glue("[{category}]({url})"),
    p = n / sum(n),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  arrange(desc(kind), -n) |>
  select(category, scope, n, p) |>
  rename_with(tools::toTitleCase) |>
  rename(`%` = p) |>
  tt() |>
  group_tt(list("Software" = 1, "Documents" = 4)) |>
  format_tt(j = 1, markdown = TRUE)
```

@@

n_platform <- sum(!is.na(oarch$platform))
p_platform <- n_platform / nrow(oarch)

@@

n_datasets <- nrow(filter(oarch, tag1 == "Datasets" | tag2 == "Datasets" | 
                            tag3 == "Datasets" | tag4 == "Datasets" | 
                            tag5 == "Datasets"))
p_datasets <- n_datasets / nrow(oarch)
```

From our breakdown of open-archaeo by category (see @tbl-categories), we can infer the pervasiveness of various development models, and the requisite technical capabilities that developers assume users hold.
Most resources (`r percent(p_platform)`) included in open-archaeo are designed to be used atop an existing "platform" -- for example a package that extends a programming language (e.g. radiocarbon calibration is implemented in R in the package 'rcarbon' or Python in the package 'iosacal') or a plugin for an application (e.g. 'ArchaeoAstroInsight' adds tools for measuring astronomical alignments to QGIS).
Essentially such projects create additional functions within the base platform that are useful for archaeological purposes.
Others create standalone software (`r percent(p_standalone)`) that can be run independently of such platforms, for example desktop or web apps.
A significant number of projects also comprise of datasets (`r percent(p_datasets)`) and non-packaged code snippets (`r percent(p_scripts)`) that have been made available for general use.

```{r tbl-platforms}
#| tbl-cap: Platforms and programming languages used by open archaeology projects
oarch |>
  drop_na(platform) |>
  count(platform) |>
  mutate(
    significant = n > 1,
    platform = if_else(significant, platform, "Other")
  ) |>
  group_by(platform) |>
  summarise(n = sum(n)) |>
  mutate(
    p = n / sum(n),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  arrange(platform == "Other", -n) |>
  rename_with(tools::toTitleCase) |>
  tt()
```
```{r data-platforms}
p_platform_r <- sum(oarch$platform == "R", na.rm = TRUE) / nrow(oarch)
p_platform_python <- sum(oarch$platform == "Python", na.rm = TRUE) / nrow(oarch)
p_platform_qgis <- sum(oarch$platform == "QGIS", na.rm = TRUE) / nrow(oarch)
```

@@

```{r tbl-tags}
#| tbl-cap: Themes of open archaeology projects. See Batist and Roe [-@batist2023 table 1] for a description of each tag's scope.
oarch |>
  pivot_longer(c(tag1, tag2, tag3, tag4, tag5), values_to = "tag",
               values_drop_na = TRUE) |>
  drop_na(tag) |>
  group_by(tag, category) |>
  summarise(n = n(), .groups = "drop_last") |>
  mutate(total = sum(n)) |>
  pivot_wider(names_from = category, values_from = n) |>
  mutate(
    tag = glue("[{tag}](https://open-archaeo.info/tags/{slugify(tag)})"),
    across(everything(), \(x) replace_na(x, 0))
  ) |>
  arrange(-total) |>
  rename(theme = tag) |>
  rename_with(tools::toTitleCase) |>
  tt() |>
  format_tt(j = 1, markdown = TRUE)
```

@@

p_license_gpl <- n_filter(oarch, str_detect(license, "GPL")) / n_filter(oarch, !is.na(license))
p_license_mit <- n_filter(oarch, str_detect(license, "MIT")) / n_filter(oarch, !is.na(license))
p_license_cc <- n_filter(oarch, str_detect(license, "CC")) / n_filter(oarch, !is.na(license))
```

### 2024-07-13 analysis/_06-network.qmd
Correct in-text stats
Also made some small wording changes, and updated the tinytable syntax for links

By contributing to shared repositories—whether with code (commits), issues, or comments—archaeologists using GitHub form a collaborative network which we can map using data from the GitHub API.
From a two-mode network with edges representing users’ contributions to repositories (Figure 6), we derived two one-mode networks that render co-contributions as connections in themselves: repositories connected by common contributors (the repository–repository graph, presented in Figure 7), and users connected by contributions to common repositories (the user–user graph, presented in Figure 8).
In both cases, number of contributions constitutes a natural measure of the strength or weight of the connection, which can be further broken down by type of contribution (commit, issue/pull request, or comment).
We identified clusters using the edge-betweenness community detection method, which operates by locating the edges that are situated along the shortest paths between all pairs of nodes, and which therefore exhibit high betweenness centrality; these edges form indirect links between otherwise completely unconnected sections of the network, and as such are the loci that distinguish groups of nodes that are more highly connected to each other than they are to others [@girvan2002].

@@

```{r tbl-repo-betweenness}
#| tbl-cap: Repositories ranked by centrality to the repository–repository network. Centrality is measured by node betweenness weighted by number of contributions.
oarch_graph_repo_repo |>
  as_tibble() |>
  arrange(-betweenness) |>
  slice(1:10) |>
  left_join(oarch, c("repo" = "gh_repo")) |>
  mutate(n_commits = map_int(gh_commits, nrow)) |>
  unite(tags, tag1:tag5, sep = "; ", na.rm = TRUE) |>
  mutate(repo = glue("[{repo}](https://github.com/{repo})")) |>
  transmute(rank = row_number(), repo, category, tags, n_commits) |>
  rename(repository = repo, commits = n_commits) |>
  rename_with(tools::toTitleCase) |>
  tt() |>
  format_tt(2, markdown = TRUE)
```

>>>>

By contributing to shared repositories—whether with code (commits), issues, or comments—archaeologists using GitHub form a collaborative network which we can map using data from the GitHub API [@fig-graph-repo-user].
Here we consider two facets of this network: repositories connected by common contributors (the repository–repository graph, @fig-graph-repo-repo), and users connected by contributions to common repositories (the user–user graph, @fig-graph-user-user).
In both cases, number of contributions constitutes a natural measure of the strength or weight of the connection, which can be further broken down by type of contribution (commit, issue/pull request, or comment).
We identified clusters using the edge-betweenness community detection method, which operates by locating the edges that are situated along the shortest paths between all pairs of nodes, and which therefore exhibit high betweenness centrality; these edges form indirect links between otherwise completely unconnected sections of the network, and as such are the loci that distinguish groups of nodes that are more highly connected to each other than they are to others [@girvan2002].

@@

```{r tbl-repo-betweenness}
#| tbl-cap: Repositories ranked by centrality to the repository–repository network. Centrality is measured by node betweenness weighted by number of contributions.
oarch_graph_repo_repo |>
  as_tibble() |>
  arrange(-betweenness) |>
  slice(1:10) |>
  left_join(oarch, c("repo" = "gh_repo")) |>
  mutate(n_commits = map_int(gh_commits, nrow)) |>
  unite(tags, tag1:tag5, sep = "; ", na.rm = TRUE) |>
  mutate(repo = glue("[{repo}](https://github.com/{repo})")) |>
  transmute(rank = row_number(), repo, category, tags, n_commits) |>
  rename(repository = repo, commits = n_commits) |>
  rename_with(tools::toTitleCase) |>
  tt() |>
  format_tt(j = 2, markdown = TRUE)
```
